STJ2012 analysis

### 2016-07-29 STJ2012_analysis.Rmd
run descriptive stats, plot clade abundances

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
```

@@

```{r data_import}
# Import OTU table
OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
                            skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=1,
                                      col.names=c("OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
phy <- phyloseq(OTU, TAX, SAM)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 4, digits = 2)
library(phyloseq); library(vegan)

taxcolors <- matrix(c(rgb(141, 211, 199, maxColorValue=255), rgb(255, 255, 179, maxColorValue=255), 
                      rgb(190, 186, 218, maxColorValue=255), rgb(251, 128, 114, maxColorValue=255), 
                      rgb(253, 180, 98, maxColorValue=255)), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF")))
```

@@

```{r data_import}
# Import OTU table
OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
                            skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=2,
                                      col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq object
phy <- phyloseq(OTU, TAX, SAM)
phy
```

## Descriptive statistics for dataset  
* Total count in OTU table: `r sum(otu_table(phy))`  
* Number of samples: `r nsamples(phy)`  
* Number of OTUs: `r ntaxa(phy)`

#### Distribution of OTU counts and reads per sample  
```{r phy_histograms, fig.height=3, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy)), plot=F)
samhist <- hist(log10(sample_sums(phy)), plot=F)

par(mfrow=c(1, 2), mar=c(3,3,1,1))
plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

* Range of OTU counts: `r range(taxa_sums(phy))`  
* Number of singletons:  `r length(taxa_sums(phy)[taxa_sums(phy) <= 1])`  
* Range of reads per sample: `r range(sample_sums(phy))`  
* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy))` ± `r sd(sample_sums(phy))`  
* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy))))` ± `r exp(sd(log(sample_sums(phy))))`

```{r reads_per_sample, fig.height=2, fig.width=8}
plot_bar(phy, x="Sample", y="Abundance")
```

## Filter dataset  
#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa <- taxa_sums(phy)[which(taxa_sums(phy) >= n)]  
# Remove taxa below threshold count
phy.f <- prune_taxa(names(taxa), phy)  
```

* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`

#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 100
# Remove samples with fewer reads than threshold
phy.f <- prune_samples(sample_sums(phy.f)>=sn, phy.f)

```

* Number of samples remaining in dataset: `r nsamples(phy.f) `

## Transform taxa count data
```{r transform}
# Convert to proportion (relative abundance)
phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
```

## Clade composition in all samples
```{r cladecomp, fig.height=6, fig.width=6}
samdat <- data.frame(sample_data(phy.f.p))
samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
typerelabund <- as.matrix(otu_table(phy.f.p)[order(data.frame(tax_table(phy.f.p))$Subtype), rownames(samdat)])
shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
shorebreaks <- which(shorebreaks==F) - 1
spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
# Make Barplot
par(mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7, mfrow=c(1, 1))
barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i",
        col=rainbow(75))#,
        #col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade])
rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
mtext(side=1, "Relative abundance", cex=0.7, line=1)
legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7,
       legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6,
       legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
# Add grouping bars for Shore
for (i in 1:length(shorebreaks)) {
  lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
  rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
}
# Add lines to separate species and species names
for (i in 1:length(spbreaks)) {
  lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
  text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
       labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
}
```

# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
#   Calculate distances to group centroids in full-dimensional principal coordinate space
#   Implemented by vegan betadisper(), based on Anderson (2006).
# Compartments/species:
```{r }
sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
anova(sambd)
boxplot(sambd)
sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))
sampt <- permutest(sambd, pairwise = T)
sampt2 <- TukeyHSD(sambd)
library(multcompView)
saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
saml2 <- multcompLetters(sampt2$group[, 4])

# Figure: Distances to centroids
# Coral species
par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
plot(sambdmeans$x[order(sambdmeans$x, decreasing=T)], type="n", ylim=c(0, 0.75),
     ylab="Distance to centroid", xaxt="n", xlab="")
arrows(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)],
       1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)],
       length=0.05, angle=90, code=3)
points(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)], cex=1, pch=21, bg="white")
text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75,
     labels=levels(SAM$Species)[order(sambdmeans$x, decreasing=T)])

text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)] + 0.03,
     labels=saml$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)] - 0.03,
     labels=saml2$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)

```

```{r strigosa}
dstr <- subset_samples(phy.f.p, Species=="strigosa")
plot_bar(dstr, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs
```
